---
title: "Interactions in R formulas"
author: "Richard J. Telford"
date: "2024-02-22"
categories: [R]
---

I happened across a tweet yesterday criticizing a paper

{{< tweet user=jfeldman_epi id=1760032277562577346 >}}

Now, while I don't have a clue what a DiD model is, I do know how to code a interaction in R and that last model most definitely has a interaction.

From the interactions with my response, it seems that many people don't know all the ways an interaction can be specified in regression models in R.

There are three ways include an interaction in a formula. 

```{r}
library(palmerpenguins)

# interaction specified explicitly with :
lm(body_mass_g ~ sex + year + sex:year, data = penguins)

# shortcut with * that expands to give the same model
lm(body_mass_g ~ sex * year, data = penguins)

# shortcut with ^2 that expands to give the same model
lm(body_mass_g ~ (sex + year)^2, data = penguins)
```

If you compare the coefficients from the three models, you will see that they are identical.

Since all three ways of coding the interaction give exactly the same result, surely one or two of these forms are redundant.

Well no. Consider a slightly more complex model where we add an extra prediction and its two - way interactions with the other variables. 


```{r}
# With :
lm(body_mass_g ~ sex + year + island + sex:year + sex:island + year:island, data = penguins)
```

Use the explicit style if you really like typing. With more than a few variables and their interactions this becomes unmanageable.


```{r}
# With *
lm(body_mass_g ~ sex * year * island, data = penguins)
```

Whoops, that went wrong. It included the three-way interaction as well. We need to remove those by subtracting it in the formula.


```{r}
# With *, two way only
lm(body_mass_g ~ sex * year * island - sex:year:island, data = penguins)
```

That's better. If we had many variables with their interactions, subtracting all the four, five or six way interactions would be really tedious.

Fortunately the last style just works.

```{r}
lm(body_mass_g ~ (sex + year + island)^2, data = penguins)
```

Concise and it gives us just what we want. If we did want the three way interactions, just replace `^2` with `^3`.

```{r}
lm(body_mass_g ~ (sex + year + island)^3, data = penguins)
```

This style is particularly useful if we want to include all available variables with the `.` shortcut. This might be the first step of a backwards selection.

```{r}
lm(body_mass_g ~ .^2, data = penguins)
```


I think all the styles of specifying interactions are useful. If you don't, blame [Wilkerson and Rogers (1973)](https://www.jstor.org/stable/2346786).
